---
title: "Analysis of NYPD Shooting Data"
author: "Mason Leonard"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Potential sources of bias in the data and my assumptions**:  
1. Over-policing of certain areas in NY can lead to more incident reports in certain areas  
2. Possible the shooting didn't cause the death, and so wasn't recorded as a death - potential for in-hospital infection or contraction of COVID etc., especially in early days with lack of PPE  
3. Lack of nearby hospital and emergency services can increase the likelihood of death from gunshot wounds and increase death statistics in certain areas, making one-to-one comparisons of geographical locations less reliable depending on the variables considered
4. My tendency to assume lots of issues are caused by, or at least influenced by, economic concerns or close proximity material concerns  

```{r, message=FALSE}
#Libraries used to run
library(knitr)
library(readr)
library(tidyverse)
library(tidyr)
library(rmarkdown)
library(lubridate)
library(readxl)
library(httr)
```

# Import New York Shooting Data
```{r}
# Import the data using read.csv
url_in  <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
shootings <- read.csv(url_in)
```

# Convert the dates to date objects and trim the table to only potentially useful columns
```{r}
# Convert dates to date objects
shootings <- shootings %>%
  mutate(OCCUR_DATE =  mdy(OCCUR_DATE))

kable(shootings[1:5, ], caption = "Shootings table with date objects.")
```

```{r}
# Remove irrelevant columns
shootings <- shootings %>%
  select(-c(LOC_OF_OCCUR_DESC, LOC_CLASSFCTN_DESC)) %>%
  select(OCCUR_DATE:VIC_RACE)

shootings_by_day <- shootings

kable(shootings[1:5, ], caption = "Cleaner shootings table without irrelevant columns.")
```

# Add in yearly BORO population numbers from Census Bureau
```{r}
url_in <- "https://www2.census.gov/programs-surveys/popest/tables/2020-2025/counties/totals/co-est2025-pop-36.xlsx"
url2_in <- "https://www2.census.gov/programs-surveys/popest/tables/2010-2020/intercensal/county/co-est2020int-pop-36.xlsx"
temp_url <- url_in
```

```{r, results=FALSE}
# import census data from .xlsx file
GET(temp_url, write_disk(tf <- tempfile(fileext = ".xlsx")))
```

```{r, message=FALSE}
boros <- read_excel(tf, 1L)
```

```{r, warning=FALSE}
# set third row values to the titles needed for the table, then set column names to row 3 and remove first 2 rows
boros[] <- lapply(boros, as.character)
boros[2] <- NULL
boros[3,1] = "BORO"
# Set cell values to date format that is recognized by mdy()
boros[3,2] = "01/01/20"
boros[3,3] = "01/01/21"
boros[3,4] = "01/01/22"
boros[3,5] = "01/01/23"
boros[3,6] = "01/01/24"
boros[3,7] = "01/01/25"
names(boros) <- boros[3,]
boros
#remove first three rows and keep only the population data where the cell value in BORO is equal to "New York"
boros <- boros[-c(1:3),]
boros <- subset(boros, BORO %in% c("New York"))

kable(boros[1:3, ], caption = "boro population estimates by year - US Census Bureau")
```

```{r, results=FALSE}
# Same process for census data from 2010 - 2020
GET(url2_in, write_disk(tf <- tempfile(fileext = ".xlsx")))
```

```{r, message=FALSE}
boros_2010 <- read_excel(tf, 1L)
```


```{r}
kable(boros_2010[1:6, ], caption = "boro population estimates by year - US Census Bureau")

boros_2010[] <- lapply(boros_2010, as.character)
boros_2010[2] <- NULL
boros_2010[3,1] = "BORO"
boros_2010[3,2] = "01/01/10"
boros_2010[3,3] = "01/01/11"
boros_2010[3,4] = "01/01/12"
boros_2010[3,5] = "01/01/13"
boros_2010[3,6] = "01/01/14"
boros_2010[3,7] = "01/01/15"
boros_2010[3,8] = "01/01/16"
boros_2010[3,9] = "01/01/17"
boros_2010[3,10] = "01/01/18"
boros_2010[3,11] = "01/01/19"
boros_2010[3,12] = "01/01/20"
names(boros_2010) <- boros_2010[3,]
boros_2010
boros_2010 <- boros_2010[-c(1:3),]

boros_2010 <- subset(boros_2010, BORO %in% c("New York"))

kable(boros_2010[1:2, ], caption = "boro population estimates by year - US Census Bureau")

boros_2010[12] <- NULL
kable(boros_2010[1:2, ], caption = "boro population estimates by year - US Census Bureau")
```

```{r}
# Merge the 2020-2025 and 2010-2020 census data into one table
boros_merge = merge(boros_2010, boros, by="BORO")
kable(boros_merge[1:2, ], caption = "boro population estimates by year - US Census Bureau")
```

```{r}
# Create pivot table to match format of shooting data: one column for date, one for population
boros_flip <- boros_merge %>%
  pivot_longer(cols = -c('BORO'),
               names_to = "OCCUR_DATE_YEAR",
               values_to = "POPULATION")
# Convert dates to date objects
boros_flip <- boros_flip %>%
  mutate(OCCUR_DATE_YEAR = mdy(OCCUR_DATE_YEAR))

kable(boros_flip[1:5, ], caption = "Boro populations")
```

# Create new table that sets dates to floor_date to categorize by month
```{r, results= "hide"}
shootings_by_month <- shootings %>% 
  mutate(OCCUR_DATE = floor_date(OCCUR_DATE, "month"))
```

# Create new table that sets dates to floor_date to categorize by year
```{r}
shootings_by_year <- shootings %>% 
  mutate(OCCUR_DATE = floor_date(OCCUR_DATE, "year"))
```

```{r}
kable(shootings_by_month[1:5, ], caption = "Shootings by month.")
kable(shootings_by_year[1:5, ], caption = "Shootings by year.")
```

# Add Population numbers to shooting data table
```{r}
# Add dates shootings_by_year so yearly population numbers can be matched to a row
monthly_added_pop <- cbind(shootings_by_month, shootings_by_year)
colnames(monthly_added_pop)[14] <- "OCCUR_DATE_YEAR"
# Remove columns to the right of OCCUR_DATE_YEAR
monthly_added_pop <- monthly_added_pop %>%
  select(OCCUR_DATE:OCCUR_DATE_YEAR)
kable(monthly_added_pop[1:5, ], caption = "Shootings by month.")
```

```{r}
# Match dates in boros_flip table to OCCUR_DATE_YEAR in monthly_added_pop table and create a POPULATION COLUMN, populated by population numbers
monthly_added_pop$POPULATION = boros_flip$POPULATION[ match(monthly_added_pop$OCCUR_DATE_YEAR, boros_flip$OCCUR_DATE_YEAR) ]
kable(monthly_added_pop[1:5, ], caption = "Shootings by month.")
```

# Graph shootings by day
```{r}
# Tally the number of shootings that happened on each day in OCCUR_DATE, then graph shootings vs date
daily_shootings <- shootings_by_day %>%
  group_by(OCCUR_DATE) %>%
  tally()
kable(daily_shootings[1:5, ], caption = "Table of Daily Shootings in NY")

daily_shootings %>%
  ggplot(aes(x = OCCUR_DATE, y = n)) +
  geom_line(aes(color = "n")) +
  geom_point(aes(color = "n")) +
  #scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "Daily Shootings in NY", y=NULL)
```
- Graph shows massive spike in shootings in the summer of 2020; potential for strain on healthcare system and increased rate of death for victims if hospitals and EMS overwhelmed during COVID

## Ten highest numbers of daily shootings
```{r}
daily_shootings %>%
  arrange(desc(n)) %>%
  slice(1:10) 
```

## Ten lowest numbers of daily shootings
```{r}
daily_shootings %>%
  arrange(n) %>%
  slice(1:10)
```

# Graph the increases and decreases in shootings by day
```{r}
# Create new table for daily increases and decreases (daily shootings minus lag)
daily_shootings_lag <- daily_shootings %>%
  mutate(daily_lag = n - lag(n))

# Graph the daily increases and decreases
daily_shootings_lag %>%
  ggplot(aes(x = OCCUR_DATE, y = daily_lag)) +
  geom_line(aes(color = "n")) +
  geom_point(aes(color = "n")) +
  #scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "Change in Daily Shootings in NY", y=NULL)
```

# Maximum value in each column for daily shootings graph
```{r}
daily_shootings[which.max(daily_shootings$n),]
```

# Maximum value in each column for daily shootings lag graph
```{r}
daily_shootings_lag[which.max(daily_shootings_lag$daily_lag),]
```

# Graph shootings by Month
```{r}
# Same process as for the daily shootings
monthly_shootings <- shootings %>%
  group_by(month = lubridate::floor_date(OCCUR_DATE, "month")) %>%
  tally()

kable(monthly_shootings[49:60, ], caption = "Table of Monthly Shootings in NY")

monthly_shootings %>%
  ggplot(aes(x = month, y = n)) +
  geom_line(aes(color = "n")) +
  geom_point(aes(color = "n")) +
  #scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "Monthly Shootings in NY", y=NULL)
```

## Ten highest numbers of monthly shootings
```{r}
monthly_shootings %>%
  arrange(desc(n)) %>%
  slice(1:10) 
```

## Ten lowest numbers of monthly shootings
```{r}
monthly_shootings %>%
  arrange(n) %>%
  slice(1:10)

summary(monthly_shootings)
```

```{r}
# create increase and decrease table by month
monthly_shootings_lag <- monthly_shootings %>%
  mutate(monthly_lag = n - lag(n))
```

## Ten highest numbers of monthly shootings, including increases
```{r}
# Show how much the shootings increased in the month with the highest number of shootings
monthly_shootings_lag %>%
  arrange(desc(n)) %>%
  slice(1:10)
```

## Ten highest numbers of monthly increases
```{r}
monthly_shootings_lag %>%
  arrange(desc(monthly_lag)) %>%
  slice(1:10)
```

# Graph the increases and decreases in shootings by Month
```{r}
monthly_shootings_lag %>%
  ggplot(aes(x = month, y = monthly_lag)) +
  geom_line(aes(color = "n")) +
  geom_point(aes(color = "n")) +
  #scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "Change in Monthly Shootings in NY", y=NULL)
```

## Combined graph of Monthly shootings and change in monthly shootings
```{r}
monthly_shootings_lag %>%
  ggplot(aes(x = month, y = n)) +
  geom_line(aes(color = "number of shootings")) +
  geom_point(aes(color = "number of shootings")) +
  geom_line(aes(y = monthly_lag, color = "monthly change")) +
  geom_point(aes(y = monthly_lag, color = "monthly change")) +
  #scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "Monthly Shootings and Monthly Change in Shootings in NY", y=NULL)
```

# Maximum and minimum values in each column for monthly shootings graph
```{r}
monthly_shootings[which.max(monthly_shootings$n),]
monthly_shootings[which.min(monthly_shootings$n),]
```

# Maximum and minimum values in each column for monthly shootings lag graph
```{r}
monthly_shootings_lag[which.max(monthly_shootings_lag$monthly_lag),]
monthly_shootings_lag[which.min(monthly_shootings_lag$monthly_lag),]
```

## Ten largest increases in monthly shootings
```{r}
monthly_shootings_lag %>%
  arrange(desc(monthly_lag)) %>%
  slice(1:10) 
```

## Ten largest decreases in monthly shootings
```{r}
monthly_shootings_lag %>%
  arrange(monthly_lag) %>%
  slice(1:10)
```

## Find the number of murders over time
```{r}
# Create function that tallies the number of true and false values (murder and non-murder) and then appends them to true and false vector
find_murders <- function(sDate){
  set_month <- shootings_by_month %>%
    filter(OCCUR_DATE == sDate)
  
  # Tally number of true and false values in STATISTICAL_MURDER_FLAG column
  set_month_tally <- set_month %>%
    group_by(STATISTICAL_MURDER_FLAG) %>%
    tally()
  
  # Create variable and set them equal to tallies
  murder_t <- set_month_tally$n[set_month_tally$STATISTICAL_MURDER_FLAG=="true"] 
  murder_f <- set_month_tally$n[set_month_tally$STATISTICAL_MURDER_FLAG=="false"]
  
  # Create numeric vector and append true and false tallies
  returns <- numeric()
  returns[1] <- murder_t
  returns[2] <- murder_f
  
  return(returns)
}
```

```{r}
# Create numeric vectors to store output tallies from find_murders function
t_vec <- numeric()
f_vec <- numeric()
```

```{r}
# Set range of dates to a vector
myDates <- seq.Date(as.Date("2006-01-01"), ("2024-12-01") , by = "month")
# Create for loop that iterates through every date and runs it through find_murders function, then appends outputs to vectors
for (i in 1:length(myDates)){
  sTable <- find_murders(myDates[i])
  t_vec[i] <- sTable[1]
  f_vec[i] <- sTable[2]
  
  }
```

```{r}
# Create new dataframe from dates and tallies
df <- data.frame(
  Date = myDates,
  Murders = t_vec,
  Non_Murders = f_vec
)
```

```{r}
# Graph the new table
df %>%
  ggplot(aes(x = Date, y = Murders)) +
  geom_line(aes(color = "Fatal")) +
  geom_point(aes(color = "Fatal")) +
  geom_line(aes(y = Non_Murders, color = "Non-Fatal")) +
  geom_point(aes(y = Non_Murders, color = "Non-Fatal")) +
  #scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "Monthly Fatal vs Non-Fatal Shootings", y=NULL)
```

```{r}
# Filter monthly table to have only months between Jan 2015 and Dec 2025
monthly_added_pop <- monthly_added_pop %>% filter(between(OCCUR_DATE, as.Date('2015-01-01'), as.Date('2025-12-31')))

kable(monthly_added_pop[1:5, ], caption = "Monthly shootings and population")
```

## Add monthly deaths per thousand and monthly shootings per thousand
```{r}
# Create new table that tallies monthly shootings using new date range
monthly_shootings <- monthly_added_pop %>%
  group_by(month = lubridate::floor_date(OCCUR_DATE, "month")) %>%
  tally()

kable(monthly_shootings[13:24, ], caption = "Table of Monthly Shootings in NY")

# Graph monthly shootings
monthly_shootings %>%
  ggplot(aes(x = month, y = n)) +
  geom_line(aes(color = "n")) +
  geom_point(aes(color = "n")) +
  #scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "Monthly Shootings in NY, 2015-2025", y=NULL)
```

```{r}
# Mutate "true" and "false" values to 0 and 1 so they can be summed
monthly_added_pop <- monthly_added_pop %>%
  mutate(STATISTICAL_MURDER_FLAG = gsub("true", 1, STATISTICAL_MURDER_FLAG)) %>%
  mutate(STATISTICAL_MURDER_FLAG = gsub("false", 0, STATISTICAL_MURDER_FLAG))


#apply(monthly_added_pop, 2, function(STATISTICAL_MURDER_FLAG) class(STATISTICAL_MURDER_FLAG))

kable(monthly_added_pop[1:5, ], caption = "test")

# Create table that adds number of deaths based on sum of STATISTICAL_MURDER_FLAG column for each month and add population values
monthly_per_mill <- monthly_added_pop %>%
  group_by(month = lubridate::floor_date(OCCUR_DATE, "month")) %>%
  summarize(deaths = sum(as.numeric(STATISTICAL_MURDER_FLAG)), population = mean(as.numeric(POPULATION))) %>%
  # Add deaths per million column
  mutate(deaths_per_mill = deaths *1000000 / population) %>%
  select(month, deaths, population, deaths_per_mill)

kable(monthly_per_mill[1:5, ], caption = "test")

# Add column for total monthly shootings by matching dates for tallied numbers from monthly_shootings table
monthly_per_mill$shootings = monthly_shootings$n[ match(monthly_per_mill$month, monthly_shootings$month) ]

# Add shootings per million column
monthly_per_mill <- monthly_per_mill %>%
  mutate(shootings_per_mill = shootings * 1000000 / population)

kable(monthly_per_mill[1:5, ], caption = "test")
```

## Model variables and plot model + predictions
```{r}
# Model deaths per million vs shootings per million
mod <- lm(deaths_per_mill ~ shootings_per_mill, data = monthly_per_mill)
summary(mod)
```

```{r}
# Find min and max values for shootings per million and deaths per million
monthly_per_mill %>% slice_min(shootings_per_mill)
monthly_per_mill %>% slice_max(shootings_per_mill)

monthly_per_mill %>%  slice_min(deaths_per_mill)
monthly_per_mill %>%  slice_max(deaths_per_mill)
```

```{r}
# Generate prediction values and add them to table
monthly_with_pred <- monthly_per_mill %>% mutate(pred = predict(mod))
kable(monthly_with_pred[1:5, ], caption = "")
```

```{r}
# Graph shootings_per_mill vs deaths_per_mill with prediction values 
monthly_with_pred %>% ggplot() +
  geom_point(aes(x = shootings_per_mill, y = deaths_per_mill), color = "blue") +
  geom_point(aes(x = shootings_per_mill, y = pred), color = "red")
```

```{r}
# Create new predition table by filtering for dates only during the lockdown
monthly_pred_only_shutdown <- monthly_with_pred %>% filter(between(month, as.Date('2020-03-01'), as.Date('2021-07-01')))
```

```{r}
# Plot same values but with lockdown dates overlaid in green
monthly_with_pred %>% ggplot() +
  geom_point(aes(x = shootings_per_mill, y = deaths_per_mill), color = "blue") +
  geom_point(aes(x = shootings_per_mill, y = pred), color = "red") +
  geom_point(data = monthly_pred_only_shutdown, aes(x = shootings_per_mill, y = deaths_per_mill), color = "green")
```

**Analysis**  
Despite increased strain on the healthcare system in New York, the percentage of deaths does not seem to increase when there was a spike in shootings in Summer 2020. It's possible that shooting victims were given priority in emergency rooms and surgery, or that the victims were sufficiently spread out between hospitals in the city that the hospitals and EMS were able to absorb the additional cases.  


```{r}
sessionInfo()
```

